D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-02-01 01intro_preliminary.tex

\subsubsection*{Contributions}
This paper presents D-Hammer, an automated tool for reasoning about labelled Dirac notation. D-Hammer uses a rich, dependently typed language to formalize labelled Dirac notation and supports common idioms for describing quantum systems, including big operators of the form \( \sum_{i \in I} a_i \ket{\phi_i}_S \) to represent indexed superpositions of states. The semantics of typable expressions are given in terms of Hilbert spaces, tailored to interpret the tensor product as an AC symbol. We leverage this interpretation to define a rich equational theory for labelled Dirac notation and prove its soundness with respect to our denotational semantics. Finally, we define an efficient normalization procedure to prove equivalence between two expressions. We then evaluate our procedure with respect to examples from the literature. Our evaluation covers examples based on plain Dirac notation, as well as examples based on
labelled Dirac notation. The main conclusions are that our approach outperforms prior work by~\cite{diracdec} and is capable of proving complex examples from the literature, including those from prior work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.

>>>>

\subsubsection*{Contributions}
This paper presents D-Hammer, an automated tool for reasoning about labelled Dirac notation. D-Hammer uses a rich, dependently typed language to formalize labelled Dirac notation and supports common idioms for describing quantum systems, including big operators of the form \( \sum_{i \in I} a_i \ket{\phi_i}_S \) to represent indexed superpositions of states. The semantics of typable expressions are given in terms of Hilbert spaces, tailored to interpret the tensor product as an AC symbol. We leverage this interpretation to define a rich equational theory for labelled Dirac notation and prove its soundness with respect to our denotational semantics. Finally, we define an efficient normalization procedure to prove equivalence between two expressions. We then evaluate our procedure with respect to examples from the literature. Our evaluation covers examples based on plain Dirac notation, as well as examples based on
labelled Dirac notation. The main conclusions are that our approach outperforms DiracDec~\cite{diracdec} to reason about plain Dirac notation and is able of proving complex examples from the literature on labelled Dirac notation, including examples from prior work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.

### 2025-02-01 02language_design.tex

\subsection{Algorithm to Decide Equivalence}
% \gb{I still think that the names first rewritings, second rewriting and de bruijn normalization are not good, in particular, de bruijn normalization is non-standard terminology that will make people familiar with the field cringe.}

The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax. We use an efficient algorithm to perform the normalization totally on $R \cup E$.

@@

As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $E$ by iterating through all possible permutations.
Our algorithm totally normalize the term, as is illustrated below.
\[
\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^E e_2' \mathop{\twoheadleftarrow}^R e_2
\qquad
\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup E} e \equiv e \mathop{\twoheadleftarrow}^{R \cup E} e_2
\]
% While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size. 
% This improvement in efficiency is also confirmed by experiments.

>>>>

\subsection{Algorithm to Prove Equivalence}
% \gb{I still think that the names first rewritings, second rewriting and de bruijn normalization are not good, in particular, de bruijn normalization is non-standard terminology that will make people familiar with the field cringe.}

The equivalence of Dirac notations are proved by normalization, which transforms equivalent expressions under a set of axioms into the same syntax. We use an efficient algorithm to perform the normalization totally on $R \cup E$.

@@

As a comparison, the previous work conducts normalization paritally on $R$, and equivalence by $E$ is proved through checking all possible permutations.
Our algorithm totally normalize the term, as is illustrated below.
\[
\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^E e_2' \mathop{\twoheadleftarrow}^R e_2
\qquad
\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup E} e \equiv e \mathop{\twoheadleftarrow}^{R \cup E} e_2
\]
% While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size. 
% This improvement in efficiency is also confirmed by experiments.

### 2025-02-01 05tool_eval.tex

\begin{figure}[h]
    \center
    % \scriptsize
    \setlength{\extrarowheight}{2pt}
    \begin{tabular}{l c c l}
        \hline
        example & source & time(s) & equation \\
        \hline
        LDN-4 & theorem & 0.03 & \( \bra{\Phi}_{(x, y)} M_y \cdot N_x \ket{\Phi}_{(x, y)} = \mathrm{tr}(M^T N) \) \\
        LDN-10 & paper~\cite{DBLP:conf/lics/ZhouBHYY21} & 5.17 & \( \mathrm{tr}_{((a',(b,b')),c')}\Big[\mathrm{tr}_{r}\Big(U_{(r,(a,b))} \cdot\Big(|s\>_r\<s|\otimes 
        \Big[V_{((a',(b,b')),c')} \cdots \)\\
        LDN-11 & paper~\cite{PALSBERG2024206} & 0.12 & \( U_{(a,b)}\cdot W_{(b,c)}\cdot V_{(a,c)} = 
        \sum_i |i\>_a\<i|\otimes \big((P_i)_c\cdot W_{(b,c)}\cdot (Q_i)_c\big)\) \\
        LDN-12 & circuit & 0.07 & \( \ket{i}_{a;b}\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} = {}_b\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} \cdot \ket{i}_{a}\) \\
        LDN-14 & circuit & 7.37 & \scriptsize{\( \textsf{CNOT}_{rq}\ket{\textrm{GHZ}}_{pqr} = (\textsf{CNOT}\ket{00})_{rq}\ket{0}_p + (\textsf{CNOT}\ket{11})_{rq}\ket{1}_p \)} \\
        LDN-18 & program & 1.01 & \( -\ket{0}_q\ket{+-}_{q_1, q_2} = X_{q_2} \ket{0}_q\ket{+-}_{q_1, q_2} \) \\
        \hline
    \end{tabular}
    \caption{Part of examples for labelled Dirac notations. See~\Cref{sec: examples for labelled} for the full list.}
    \label{fig: labelled examples}
\end{figure}

>>>>

\begin{figure}[h]
    \center
    % \scriptsize
    \setlength{\extrarowheight}{2pt}
    \begin{tabular}{l c c l}
        \hline
        example & source & time(s) & equation \\
        \hline
        LDN-4 & theorem & 0.03 & \( M_{r_1}\sum_{i}\ket{(i,i)}_{(r_1,r_2)} = M^T_{r_2}\sum_{i}\ket{(i,i)}_{(r_1, r_2)} \)\\
        LDN-10 & paper~\cite{DBLP:conf/lics/ZhouBHYY21} & 5.17 & \( \mathrm{tr}_{((a',(b,b')),c')}\Big[\mathrm{tr}_{r}\Big(U_{(r,(a,b))} \cdot\Big(|s\>_r\<s|\otimes 
        \Big[V_{((a',(b,b')),c')} \cdots \)\\
        LDN-11 & paper~\cite{PALSBERG2024206} & 0.12 & \( U_{(a,b)}\cdot W_{(b,c)}\cdot V_{(a,c)} = 
        \sum_i |i\>_a\<i|\otimes \big((P_i)_c\cdot W_{(b,c)}\cdot (Q_i)_c\big)\) \\
        LDN-12 & circuit & 0.07 & \( \ket{i}_{a;b}\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} = {}_b\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} \cdot \ket{i}_{a}\) \\
        LDN-14 & circuit & 7.37 & \scriptsize{\( \textsf{CNOT}_{rq}\ket{\textrm{GHZ}}_{pqr} = (\textsf{CNOT}\ket{00})_{rq}\ket{0}_p + (\textsf{CNOT}\ket{11})_{rq}\ket{1}_p \)} \\
        LDN-16 & circuit & 0.08 & \scriptsize{\( M_{prq}\ket{\textrm{GHZ}}_{prq} = N_{rq} \ket{\textrm{GHZ}}_{pqr}, M \triangleq \sum_{ij}\ket{ij}\bra{ij}\otimes U_{ij} \cdots \)} \\
        \hline
    \end{tabular}
    \caption{Part of examples for labelled Dirac notations. See~\Cref{sec: examples for labelled} for the full list.}
    \label{fig: labelled examples}
\end{figure}

### 2025-02-01 main.tex
add figure
update section 3

\usepackage{wasysym}
\usepackage{extarrows}
\usepackage{tikz}

@@

@@

\newcommand {\cH } {{\mathcal{H}}}
\newcommand {\bu } {{\mathbf{u}}}
\newcommand {\bv } {{\mathbf{v}}}
\newcommand {\bC } {{\mathbb{C}}}
\newcommand {\ghz } {{\mathrm{GHZ}}}
\newcommand {\cnot } {{\mathsf{CNOT}}}
\newcommand {\ccnot } {{\mathsf{CCNOT}}}
\newcommand {\swap } {{\mathsf{SWAP}}}
\renewcommand{\matrix}[1]{\begin{bmatrix}#1\end{bmatrix}}

@@

%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Labelled Dirac notation is a formalism commonly used by physicists to
represent many-body quantum systems and computer scientists to assert
properties of quantum programs. Labelled Dirac notation is supported
by a rich equational theory for proving equality between expressions
of the language. These proofs are typically carried on pen-and-paper,
and can be exceedingly long and error-prone. We introduce D-Hammer,
the first tool to supported automated equational proof for labelled
Dirac notation. The salient features of D-Hammer are: an expressive,
higher-order, dependently-typed language for labelled Dirac notation;
an efficient algorithm for normalization; and an optimized
\CC\ implementation. We evaluate the implementation on representative
examples from plain and labelled Dirac notation. In the case of plain
Dirac notation, we show that our implementation significantly
outperforms DiracDec (Xu et al, POPL'25).

>>>>

\usepackage{wasysym}
\usepackage{extarrows}
\usepackage{tikz}
\usetikzlibrary{tikzmark}
\usepackage{annotate-equations}

@@

\usepackage{tikz}
\usetikzlibrary{quantikz2}

@@

\newcommand {\cH } {{\mathcal{H}}}
\newcommand {\bu } {{\mathbf{u}}}
\newcommand {\bv } {{\mathbf{v}}}
\newcommand {\bC } {{\mathbb{C}}}
\newcommand {\ghz } {{\mathrm{GHZ}}}
\newcommand {\cnot } {{\mathsf{CNOT}}}
\newcommand {\ccnot } {{\mathsf{CCNOT}}}
\newcommand {\Swap } {{\mathsf{SWAP}}}
\renewcommand{\matrix}[1]{\begin{bmatrix}#1\end{bmatrix}}

@@

%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    Labelled Dirac notation is a formalism commonly used by physicists to represent many-body quantum systems and by computer scientists to assert properties of quantum programs. It is supported by a rich equational theory for proving equality between expressions in the language. These proofs are typically carried on pen-and-paper, and can be exceedingly long and error-prone. We introduce D-Hammer, the first tool to support automated equational proof for labelled Dirac notation. The salient features of D-Hammer include: an expressive, higher-order, dependently-typed language for labelled Dirac notation; an efficient normalization algorithm; and an optimized \CC\ implementation. We evaluate the implementation on representative examples from both plain and labelled Dirac notation. In the case of plain Dirac notation, we show that our implementation significantly outperforms DiracDec (Xu et al., POPL'25).
